Case based Arabic Morphological Analysis with Compositional Non-deterministic Automata

### 2010-06-26 tex/sarf.tex
spell

We hypothesize that many NLP case studies need the 
morphological analyzer to answer simple queries that do not need 
high complexity and accuracy at the low morphological analysis level 
and that can often compensate for tolerable inaccuracy at higher levels. 
For example, if the query concerns proper names and the 
prefix in question in the analysis can only connect to a verb, 
we can provide an answer without completing the analysis. 
%TODO A lso if a word such as \RL{wasiim} was name was ambiguous and we were looking for a sequence of names, then 
We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
with features demanding resolution at an abstraction level
higher than the text itself
led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
We strongly believe and we find evidence in our results that a 
case-based morphological analyzer that allows a case-based
controller to intervene at every decision to 
guide, use, prune, and refine the morphological analysis
is of high utility.

@@

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely as long 
as the text is valid.
The controller is also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the sun of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of a three level deep
complex structure. 
We also showcase that the results of the accurate high level abstraction
results of the case-based analysis was not affected by a major
refinement to the analyzer that increased low level analysis accuracy.

@@

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
\item {\bf Exhaustive analysis:} Since we have a case-based 
controller that can decide on the fly whether an analysis is 
accepted or not, we can afford to keep all valid analyses modulo
those pruned by the case controller. 
We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
This allows Sarf to perform better and 
better retain compositional part of speech
tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

The rest of this paper is structured as follows. In Section~\ref{sec:related}
we compare Sarf to related work. Then in Section~\ref{sec:sarf}
we discuss Sarf, illustrate the way it works with a running example
that also explains how non-deterministic finite state automata
works. In Section~\ref{sec:islamic} we present the hadith case
study. We present our results in Section~\ref{sec:results}
and we conclude with future work in Section~\ref{sec:future}.

@@

The survey work in~\cite{Sughaiyer:04} discusses and compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the morphological 
analyzer itself or use a specific set of part of speech tags
as their reference.
Sarf differs in that it is a general and rich morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller that drives
the morphological analyzer prunes these decisions. 
Up to our knowledge, Sarf is the only morphological analyzer
that addresses the ``run-on'' words problem.

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with a non-deterministic composition of three
FSAs that are the prefix, stem, and suffix FSAs. 
With Sarf it is easy to control the FSAs with a custom user case 
controller, while with Beesley and up to our understanding the 
effort requires recompiling the Xerox rules.

@@

We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
is the closest to our approach. SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. We find supporting evidence in this decision to our hypothesis
about the utility of a case-based analyzer. 
The refinements that resulted in algorithmic changes in SAMA were 
done manually and worked in integration with the ATB format. 
In Sarf we propose a case-based controller approach that can modify 
the behavior of the analyzer on the fly and that should interact
with the case study. In Sarf the analyzer remains general, and the 
case-based controller specializes the analysis.

@@

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries in our analysis 
to run the prefix, suffix and stem FSAs. 
We think that the inferential-realizational approach 
of ElixierFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammar correct.
We can compare the case-based controller in Sarf to a high level
component in the ElixirFM domain specific language.

@@

Finally, The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an Automata that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others.

@@

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
  Our understanding is that those books are historical documents with 
  public IP.}.

@@

The FSA reports a valid chain of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the begining of the hadith with the begining of the sequence,
and marks the end of the hadith by the begining of the next sequence. 
It also marks the chain of narrators as the sequence itself.

@@

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
 & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 24.5 & 22.67 & & 23.5 & 23.17 & & 2.25 & 1.67\\ \hline
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

To evaluate our hypothesis of the utility of a case-based morphological analyzer,
we ran our case study in two modes, simple and accurate.
In the simple mode Sarf is controlled by the hadith controller.
In the accurate mode we apply several refinements to Sarf.
The first refinement removes female proper names since they do not occur often in
hadith.
We also add a refinement that detects compound names as they are common in narrator
names.
Finally, we consider a possessive form as a reference to a person and treat
it as a narrator.

We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
The word count row is an indicator of the size of data.
The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
narrators, and a narrator contains names.
In general, we observe that the refinements slightly improved performance and accuracy at
the lowest level of abstraction by reducing false positives.
However, the refienements had little to no effect on the high accuracy measures at higher
levels of abstraction.
We beleive that this is a strong supporting evidence to our hypothesis.

The names row reports the total number of detected names. The names/narator row reports the average
number of names per narrator structure. The narrator row reports the number of narrators detected.
The narrator/chain reports the number of narrators per chain structure. And the chain row reports
the number of detected chains in the book.
The ignored names row reports the number of names detected outside the chains and thus irrelevant to
the case study.
The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
The narrator name accuracy row reflects the percentage of correct proper names detected within
narrator strutures.

The name false positives row reports the percentage of words recognized as names while they should not.
We notice that the low level metric of name false positives improved dramatically with the accurate
analysis without substantially affecting the high abstract measures such as chain and segmentation
accuracy.
We notice that the controller in the simple analysis ignored more names and thus the case-based
controller compensated for the inaccuracy of the low level analysis with a simple heuristic and
without the need of the fancy refinements.

We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than the accurate
analysis and discovered that the chains detected by the accurate analysis were a subset of those detected by the simple
analysis. 
This means that the simple analysis results have false positives and did not miss important information. We hypothesize
that a higher level of abstraction analysis such as intersecting the narrators with their biographies can 
reduce false positives in the simple analysis.

@@

In the future we plan to complete the hadith case study 
and automate narrator discovery and authentication in the 
biography books. 
We plan to make use of the name connector and narrator
connector stop phrases that our case study extracted to create a 
learning system that can learn more names. 
We also plan on developing a domain specific language similar 
that allows users to easily specify their case-based controller
to interact with Sarf.

>>>>

We hypothesize that many NLP case studies need the 
morphological analyzer to answer simple queries that do not need 
high complexity and accuracy at the low morphological analysis level 
and that can often compensate at higher level for tolerable inaccuracy at low levels. 
For example, if a query concerns proper names and the 
prefix in question in the analysis connects only to a verb, 
we do not need to further analyze the rest of the word 
to provide an answer.
We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
with features demanding resolution at an abstraction level
higher than the text itself
led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
We strongly believe and we find evidence in our results that a 
case-based morphological analyzer that allows a case-based
controller to intervene at every decision to 
guide, use, prune, and refine the morphological analysis
is of high utility.

@@

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely as long 
as the text is valid.
The controller is also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of a three level deep
complex structure. 
We also showcase that the results of the accurate high level abstraction
results of the case-based analysis was not affected by a major
refinement to the analyzer that increased low level analysis accuracy.

@@

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
\item {\bf Exhaustive analysis:} Since we have a case-based 
controller that can decide on the fly whether an analysis is 
accepted or not, we can afford to keep all valid analyses modulo
those pruned by the case controller. 
We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
This allows Sarf to perform better and 
maintain compositional part of speech tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

The rest of this paper is structured as follows. In Section~\ref{sec:related}
we compare Sarf to related work. Then in Section~\ref{sec:sarf}
we discuss Sarf and illustrate the way it works with a running example.
We also explain how non-deterministic finite state Automata
work. In Section~\ref{sec:islamic} we present the hadith case
study. We present our results in Section~\ref{sec:results}
and conclude with future work in Section~\ref{sec:future}.

@@

The survey in~\cite{Sughaiyer:04} discusses and compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the morphological 
analyzer itself or use a specific set of part of speech tags
as their reference.
Sarf differs in that it is a general and rich morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller that drives
the morphological analyzer prunes these decisions.

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with a non-deterministic composition of three
FSAs that are the prefix, stem, and suffix FSAs. 
With Sarf it is easy to control the FSAs with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

@@

We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
is the closest to our approach. SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. We find supporting evidence in this decision to our hypothesis
about the utility of a case-based analyzer. 
The refinements that resulted in algorithmic changes in SAMA were 
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with the any case study. In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

@@

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries in our analysis 
to run the prefix, suffix and stem FSAs. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammatically correct.
%We can compare the case-based controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

Finally, the case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an Automata that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others.

@@

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

The FSA reports a valid chain of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of the sequence,
and marks the end of the hadith by the beginning of the next sequence. 
It also marks the chain of narrators as the sequence itself.

@@

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
 & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\ \hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 24.5 & 22.67 & & 23.5 & 23.17 & & 2.25 & 1.67\\ \hline
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

To evaluate our hypothesis of the utility of a case-based morphological analyzer,
we ran our case study in two modes, simple and refined.
In the simple mode, Sarf is controlled by the hadith controller.
In the refined mode, we apply several refinements to Sarf.
The first refinement removes female proper names since they do not occur often in
hadith.
We also add a refinement that detects compound names as they are common in narrator
names.
Finally, we consider a possessive form as a reference to a person and treat
it as a narrator.

We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
The word count row is an indicator of the size of data.
The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
narrators, and a narrator contains names.
In general, we observe that the refinements slightly improved performance and accuracy at
the lowest level of abstraction by reducing false positives.
However, the refinements had little to no effect on the high accuracy measures at higher
levels of abstraction.
We believe that this is a strong supporting evidence to our hypothesis.

The names row reports the total number of detected names. The names/narrator row reports the average
number of names per narrator structure. The narrators row reports the number of narrators detected.
The narrators/chain row reports the number of narrators per chain structure. And the chain row reports
the number of detected chains in the book.
The ignored names row reports the number of names detected outside the chains and thus irrelevant to
the case study.
The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
The narrator name accuracy row reflects the percentage of correct proper names detected within
narrator structures.

The name false positives row reports the percentage of words recognized as names while they should not.
We notice that the low level metric of name false positives improved dramatically with the refined
analysis without substantially affecting the high abstract measures such as chain and segmentation
accuracy.
We notice that the controller in the simple analysis ignored more names and thus the case-based
controller compensated for the inaccuracy of the low level analysis with a simple heuristic and
without the need of the fancy refinements.

We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than that of the refined
analysis. We  discovered that the chains detected by the refined analysis were a subset of those detected by the simple
analysis. 
The reason for the lower accuracy is false positive chains. 
The simple analysis did not miss important information. 
We hypothesize that a higher level of abstraction such as intersecting the narrators from the hadith books with the narrators 
in the biographies can reduce the false positives in the chains.

@@

In the future, we plan to complete the hadith case study 
and automate narrator discovery and authentication in the 
biography books. 
We plan to make use of the name connector and narrator
connector stop phrases that our case study extracted to create a 
learning system that can learn more names. 
We also plan on developing a domain specific language similar 
to ElexirFM~\cite{Otakar:07}
that allows users to easily specify their case-based controller
to interact with Sarf.
